Automatically Splitting a Two-Stage Lambda Calculus

### 2015-10-12 paper/examples.tex
Pound on program bifurcation really hard in the related work.
format fiddling
lots of cleaning and compression

1`dot : vec * vec -> $int
fun dot (
  (3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
  (3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
next{
  2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
1`}`
\end{lstlisting}
%
The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

Our algorithm splits \texttt{dot} into the two functions:
%
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()),(x2,y2,())) 
  = ((), (x1*x2)+(y1*y2))`

2`fun dot2 ((((),(),z1),((),(),z2)),l) 
  = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, except
that they write the precomputation into a mutable cache.

@@

Stage-splitting \texttt{qss} produces:
%
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
datatype tree = Leaf
              | Branch of int * tree * int * tree

1`fun partition ((p,l):int*list) : (int*list*list) = 
  case l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) => 
      let val (s,left,right) = partition (p,t) in 
      if h<p 
      then (s+1,Cons(h,left),right) 
      else (s,left,Cons(h,right))

fun qs1 (l : list, k : unit) = 
  ((), case l of
    Empty => Leaf
  | Cons (h,t) => 
      Branch (
        let val (left,right,n) = partition (h,t) in
        (n, #2 (qs1 left k), h, #2 (qs1 right k))))`

@@

To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
the second pass, evaluate \texttt{qs2} on that tree and each rank.
\begin{lstlisting}
1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
\end{lstlisting}
%
In the appendix provided as supplemental material, we extend this example by
splitting \texttt{map (qss l)} over an arbitrary list of ranks \texttt{ks}. We also
provide an example of a program whose boundary data structure is a trie.

\subsection {Composing Graphics Pipeline Programs}

>>>>

1`dot : vec * vec -> $int
fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
  next{ 2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)` 1`}`
\end{lstlisting}
%
The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

Our algorithm splits \texttt{dot} into the two functions:
%
\begin{lstlisting} 
1`fun dot1 ((x1,y1,()), (x2,y2,()))    = ((), (x1*x2)+(y1*y2))`
2`fun dot2 ((((),(),z1),((),(),z2)),l) = l+(z1*z2)`
\end{lstlisting}
%
As expected, \texttt{dot1} returns \texttt{(x1*x2)+(y1*y2)} as the
precomputation, and \texttt{dot2} adds that precomputation to the products of
the final coordinates. This is exactly what is done in \cite{knoblock96}, except
that they write the precomputation into a mutable cache.

@@

Stage-splitting \texttt{qss} produces:
%
\begin{lstlisting} 
datatype tree = Leaf
              | Branch of int * tree * int * tree
1`datatype list = Empty | Cons of int * list

fun part ((p,l):int*list) : (int*list*list) = 
  case l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) => 
      let val (n,left,right) = part (p,t) in 
      if h<p 
      then (n+1,Cons(h,left),right) 
      else (n,left,Cons(h,right))

fun qs1 (l : list, k : unit) = 
  ((), case l of
    Empty => Leaf
  | Cons (h,t) => 
      Branch (
        let val (n,left,right) = part (h,t) in
        (n, #2 (qs1 left k), h, #2 (qs1 right k))))`

@@

To run many order statistics queries $\texttt{k1},\dots,\texttt{km}$ on the same
list, use \texttt{qs1} to compute the boundary tree in the first pass, and in
the second pass, evaluate \texttt{qs2} on that tree and each rank.
\begin{lstlisting}
1`let val (_,t) = qs1 ([7,4,2,5,9,...,3],())
in `2`(qs2 (((),k1),1`t`),...,qs2 (((),km),1`t`))`
\end{lstlisting}
%

\subsection{Composing Graphics Pipeline Programs}
\label{sec:graphics}

### 2015-10-12 paper/overview.tex
lots of cleaning and compression

\begin{theorem}
  Consider an execution of \texttt{qs1} with a randomly permuted input
  of $n$ keys and performing $m$ executions of \texttt{qs2} with the
  result of \texttt{qs1}.  The total run-time for this computation is
  $\Theta(n\log{n} + m\log{n})$ in expectation.
\end{theorem}
\begin{proof}
  We note first that \texttt{qs1} is
  isomorphic to a quicksort algorithm but builds a binary search tree
  instead of combining the returned results in a list.  The run-time
  for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
%
  The depth of the tree generated by \texttt{qs1} is isomorphic to the
    depth of a run of the quickselect algorithm, or equivalently to the
    span of a parallel implementation of the quicksort algorithm, both
    of which are known to be (expected) $\Theta(\log{n})$.
%
    By inspection of \texttt{qs2}, we can see that the algorithm takes
    time proportional to the depth of the binary search tree, which is
    $\Theta(\log{n})$. Consider now a random variable for each of the
    $m$ invocations of \texttt{qs2}, each of which takes expected
    $\Theta(\log{n})$ time.  Since the total time is the sum of these
    random variables, and the expectation of the sum is the sum of
    the expectations of the random variables, the bound follows.
\end{proof}

>>>>

% \begin{theorem}
%   Consider an execution of \texttt{qs1} with a randomly permuted input
%   of $n$ keys and performing $m$ executions of \texttt{qs2} with the
%   result of \texttt{qs1}.  The total run-time for this computation is
%   $\Theta(n\log{n} + m\log{n})$ in expectation.
% \end{theorem}
% \begin{proof}
%   We note first that \texttt{qs1} is
%   isomorphic to a quicksort algorithm but builds a binary search tree
%   instead of combining the returned results in a list.  The run-time
%   for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
% %
%   The depth of the tree generated by \texttt{qs1} is isomorphic to the
%     depth of a run of the quickselect algorithm, or equivalently to the
%     span of a parallel implementation of the quicksort algorithm, both
%     of which are known to be (expected) $\Theta(\log{n})$.
% %
%     By inspection of \texttt{qs2}, we can see that the algorithm takes
%     time proportional to the depth of the binary search tree, which is
%     $\Theta(\log{n})$. Consider now a random variable for each of the
%     $m$ invocations of \texttt{qs2}, each of which takes expected
%     $\Theta(\log{n})$ time.  Since the total time is the sum of these
%     random variables, and the expectation of the sum is the sum of
%     the expectations of the random variables, the bound follows.
% \end{proof}

### 2015-10-12 paper/semantics.tex
lots of cleaning and compression

All three worlds contain unit, product, and function types defined in the usual fashion;
sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
\footnote{We omit sums and recursive types from \bbonem\ to save space, though we do not believe they would present any unique difficulty.}
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.

@@

\input{figures/dynamics}
\input{figures/dynamics-overview}

@@

>>>>

All three worlds contain unit, product, function, sum and recursive types defined in the usual fashion.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.

@@

\input{figures/dynamics}

@@

\begin{abstrsyn}
\begin{theorem} [Progress] \leavevmode
\label{thm:progress} 
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem} [Preservation] \leavevmode
\label{thm:preservation} 
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}
\end{abstrsyn}

### 2015-10-12 paper/splitting.tex
lots of cleaning and compression

The goal of a stage splitting translation 
is to send a program $\mathcal P$ in a multistage language
to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
By equivalent, we mean that $\mathcal P$ and $\mathcal P'$ should produce 
the same answer under their respective semantics.

@@

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
Since the rules are simply recursive on the structure of the term, 
the splitting algorithm runs in linear time on the size of the input program.
Splitting is defined for all well-typed inputs (\ref{thm:total} in \ref{fig:splittingSummary}), and it produces unique results (\ref{thm:unique}).
That is, each splitting judgement defines a total function.
We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

@@

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
From here, the inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
To close this gap we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.
After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.

In order to prove \ref{lem:twoCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds of multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

@@

\paragraph{Reduction of Application.} Let $E = \app {\exv{\fix f x e}} {\exv v}$,
$E' = [\fix fxe, v/f,x] e$, \\ 
$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$, \\
$Q = \fix f {\tup{x,\roll l}} r$, and \\
$C = \left[
\talllet{\tup{y_1,z_1}}{\exv {\tup {I,\tup{}}}}{
\talllet{\tup{y_2,z_2}} {\exv {\tup {i_1,\tup{}}}}{
\talllet{\tup{y_3,z_3}}{\app{\exv {y_1}}{\exv {y_2}}}{\exv{\tup{y_3,\tup{z_1,z_2,z_3}}}\ttrpar\ttrpar\ttrpar}}}\right]$.
\\
We are given $\stepmix E {E'}$,
$\splitone E A C {\tup{\_,\_,l}.\app{\exv Q}{\exv{\tup{q_1,l}}}}$,
$\splitone {E'} A {[I,i_1/f,x]c} {l.[Q,q_1/f,x]r}$,
and $\reduce {[I,i_1/f,x]c} {\tup{i,b}}$.
From this, we can directly derive $\reduce C {\tup{i,\tup {\tup{},\tup{},\roll b}}}$
and $\app{\exv Q}{\exv{\tup{q_1,\roll b}}} \equiv [Q,q_1,b/f,x,l] r$.
This pattern applies to all of the other reduction rules involving substitution,
namely those for {\tt caseg}, and {\tt letg}. 
Observe that, in our formal version of the language, 
{\tt caseg} is the only construct where information is added to the boundary value.

>>>>

The goal of a stage splitting translation 
is to send a program $\mathcal P$ in a multistage language
to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
More specifically, $\mathcal P$ and $\mathcal P'$ should produce 
the same answer under their respective semantics.

@@

The rules of the three splitting judgements are given in \ref{fig:termSplit}.
Since the rules are simply recursive on the structure of the term, 
the splitting algorithm runs in linear time on the size of the input program.
Splitting is defined for all well-typed inputs (\ref{thm:total}), and it produces unique results (\ref{thm:unique}).
That is, each splitting judgement defines a total function.

\begin{theorem} [Splitting Totality] \leavevmode
\label{thm:total} 
\begin{itemize} 
\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
\item For term $e$, if $\typesone e A$, then $\splitone e A c {l.r}$.
\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
\end{itemize}
\end{theorem}

\begin{theorem} [Splitting Uniqueness] \leavevmode
\label{thm:unique}
\begin{itemize} 
\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, and $l.r = l'.r'$.
\item If $\splitone e A p {l.r}$ and $\splitone e A {c'} {l'.r'}$, then $c = c'$, and $l.r = l'.r'$.
\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
\end{itemize}
\end{theorem}

We prove these theorems by straightforward induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

@@

\begin{theorem} [End-to-End Correctness]
\label{thm:allCorrect}
If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}

@@

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
From here, the inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
To close this gap we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.

\begin{lemma} [Single Step Correctness]
\label{lem:twoCorrect}
\begin{itemize} 
\item If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\item If $e \stepsym \bbonem e'$, $\splitone e A c {l.r}$, $\splitone {e'} A {c'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{itemize}
\end{lemma}

After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.
In order to prove \ref{lem:twoCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds of multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

\begin{lemma} [Hoisting Step Correctness]
\label{lem:hoistCorrect}
If $\lift e {e'}$, $\splitone e A c {l.r}$, and $\splitone {e'} A {c'} {l'.r'}$, then $c = c'$, and $l.r \equiv l'.r'$.
\end{lemma}

@@

\paragraph{Reduction of Application.} Let $E = \app {\exv{\fix f x e}} {\exv v}$,
$E' = [\fix fxe, v/f,x] e$, \\ 
$I = \fix f x {\letin {\tup{x,y}} c {\tup{x,\roll y}}}$,
$Q = \fix f {\tup{x,\roll l}} r$, and \\
$C = \left[
\talllet{\tup{y_1,z_1}}{\exv {\tup {I,\tup{}}}}{
\talllet{\tup{y_2,z_2}} {\exv {\tup {i_1,\tup{}}}}{
\talllet{\tup{y_3,z_3}}{\app{\exv {y_1}}{\exv {y_2}}}{\exv{\tup{y_3,\tup{z_1,z_2,z_3}}}\ttrpar\ttrpar\ttrpar}}}\right]$.
\\
We are given $\stepmix E {E'}$,
$\splitone E A C {\tup{\_,\_,l}.\app{\exv Q}{\exv{\tup{q_1,l}}}}$,
$\splitone {E'} A {[I,i_1/f,x]c} {l.[Q,q_1/f,x]r}$,
and $\reduce {[I,i_1/f,x]c} {\tup{i,b}}$.
From this, we can directly derive $\reduce C {\tup{i,\tup {\tup{},\tup{},\roll b}}}$
and $\app{\exv Q}{\exv{\tup{q_1,\roll b}}} \equiv [Q,q_1,b/f,x,l] r$.
This pattern applies to all of the other reduction rules involving substitution,
namely those for {\tt caseg}, and {\tt letg}. 
Observe that, in our formal version of the language, 
{\tt caseg} is the only construct where information is added to the boundary value.
